\documentclass[letter-paper]{tufte-book}

%%
% Book metadata
\title{Statistical Mechanics}
\author[]{Inusuke Shibemoto}
%\publisher{Research Institute of Valinor}

%%
% If they're installed, use Bergamo and Chantilly from www.fontsite.com.
% They're clones of Bembo and Gill Sans, respectively.
\IfFileExists{bergamo.sty}{\usepackage[osf]{bergamo}}{}% Bembo
\IfFileExists{chantill.sty}{\usepackage{chantill}}{}% Gill Sans

%\usepackage{microtype}
\usepackage{amssymb}
\usepackage{amsmath}
%%
% For nicely typeset tabular material
\usepackage{booktabs}

%% overunder braces
\usepackage{oubraces}

%% 
\usepackage{xcolor}
\usepackage{tcolorbox}

\newtcolorbox[auto counter,number within=section]{derivbox}[2][]{colback=TealBlue!5!white,colframe=TealBlue,title=Box \thetcbcounter:\ #2,#1}                                                          

\makeatletter
\@openrightfalse
\makeatother

%%
% For graphics / images
\usepackage{graphicx}
\setkeys{Gin}{width=\linewidth,totalheight=\textheight,keepaspectratio}
\graphicspath{{figs/}}

% The fancyvrb package lets us customize the formatting of verbatim
% environments.  We use a slightly smaller font.
\usepackage{fancyvrb}
\fvset{fontsize=\normalsize}

\usepackage[plain]{fancyref}
\newcommand*{\fancyrefboxlabelprefix}{box}
\fancyrefaddcaptions{english}{%
  \providecommand*{\frefboxname}{Box}%
  \providecommand*{\Frefboxname}{Box}%
}
\frefformat{plain}{\fancyrefboxlabelprefix}{\frefboxname\fancyrefdefaultspacing#1}
\Frefformat{plain}{\fancyrefboxlabelprefix}{\Frefboxname\fancyrefdefaultspacing#1}

%%
% Prints argument within hanging parentheses (i.e., parentheses that take
% up no horizontal space).  Useful in tabular environments.
\newcommand{\hangp}[1]{\makebox[0pt][r]{(}#1\makebox[0pt][l]{)}}

%% 
% Prints an asterisk that takes up no horizontal space.
% Useful in tabular environments.
\newcommand{\hangstar}{\makebox[0pt][l]{*}}

%%
% Prints a trailing space in a smart way.
\usepackage{xspace}
\usepackage{xstring}

%%
% Some shortcuts for Tufte's book titles.  The lowercase commands will
% produce the initials of the book title in italics.  The all-caps commands
% will print out the full title of the book in italics.
\newcommand{\vdqi}{\textit{VDQI}\xspace}
\newcommand{\ei}{\textit{EI}\xspace}
\newcommand{\ve}{\textit{VE}\xspace}
\newcommand{\be}{\textit{BE}\xspace}
\newcommand{\VDQI}{\textit{The Visual Display of Quantitative Information}\xspace}
\newcommand{\EI}{\textit{Envisioning Information}\xspace}
\newcommand{\VE}{\textit{Visual Explanations}\xspace}
\newcommand{\BE}{\textit{Beautiful Evidence}\xspace}

\newcommand{\TL}{Tufte-\LaTeX\xspace}

% Prints the month name (e.g., January) and the year (e.g., 2008)
\newcommand{\monthyear}{%
  \ifcase\month\or January\or February\or March\or April\or May\or June\or
  July\or August\or September\or October\or November\or
  December\fi\space\number\year
}


\newcommand{\urlwhitespacereplace}[1]{\StrSubstitute{#1}{ }{_}[\wpLink]}

\newcommand{\wikipedialink}[1]{http://en.wikipedia.org/wiki/#1}% needs \wpLink now

\newcommand{\anonymouswikipedialink}[1]{\urlwhitespacereplace{#1}\href{\wikipedialink{\wpLink}}{Wikipedia}}

\newcommand{\Wikiref}[1]{\urlwhitespacereplace{#1}\href{\wikipedialink{\wpLink}}{#1}}

% Prints an epigraph and speaker in sans serif, all-caps type.
\newcommand{\openepigraph}[2]{%
  %\sffamily\fontsize{14}{16}\selectfont
  \begin{fullwidth}
  \sffamily\large
  \begin{doublespace}
  \noindent\allcaps{#1}\\% epigraph
  \noindent\allcaps{#2}% author
  \end{doublespace}
  \end{fullwidth}
}

% Inserts a blank page
\newcommand{\blankpage}{\newpage\hbox{}\thispagestyle{empty}\newpage}

\usepackage{units}

% Typesets the font size, leading, and measure in the form of 10/12x26 pc.
\newcommand{\measure}[3]{#1/#2$\times$\unit[#3]{pc}}

% Macros for typesetting the documentation
\newcommand{\hlred}[1]{\textcolor{Maroon}{#1}}% prints in red
\newcommand{\hangleft}[1]{\makebox[0pt][r]{#1}}
\newcommand{\hairsp}{\hspace{1pt}}% hair space
\newcommand{\hquad}{\hskip0.5em\relax}% half quad space
\newcommand{\TODO}{\textcolor{red}{\bf TODO!}\xspace}
\newcommand{\na}{\quad--}% used in tables for N/A cells
\providecommand{\XeLaTeX}{X\lower.5ex\hbox{\kern-0.15em\reflectbox{E}}\kern-0.1em\LaTeX}
\newcommand{\tXeLaTeX}{\XeLaTeX\index{XeLaTeX@\protect\XeLaTeX}}
% \index{\texttt{\textbackslash xyz}@\hangleft{\texttt{\textbackslash}}\texttt{xyz}}
\newcommand{\tuftebs}{\symbol{'134}}% a backslash in tt type in OT1/T1
\newcommand{\doccmdnoindex}[2][]{\texttt{\tuftebs#2}}% command name -- adds backslash automatically (and doesn't add cmd to the index)
\newcommand{\doccmddef}[2][]{%
  \hlred{\texttt{\tuftebs#2}}\label{cmd:#2}%
  \ifthenelse{\isempty{#1}}%
    {% add the command to the index
      \index{#2 command@\protect\hangleft{\texttt{\tuftebs}}\texttt{#2}}% command name
    }%
    {% add the command and package to the index
      \index{#2 command@\protect\hangleft{\texttt{\tuftebs}}\texttt{#2} (\texttt{#1} package)}% command name
      \index{#1 package@\texttt{#1} package}\index{packages!#1@\texttt{#1}}% package name
    }%
}% command name -- adds backslash automatically
\newcommand{\doccmd}[2][]{%
  \texttt{\tuftebs#2}%
  \ifthenelse{\isempty{#1}}%
    {% add the command to the index
      \index{#2 command@\protect\hangleft{\texttt{\tuftebs}}\texttt{#2}}% command name
    }%
    {% add the command and package to the index
      \index{#2 command@\protect\hangleft{\texttt{\tuftebs}}\texttt{#2} (\texttt{#1} package)}% command name
      \index{#1 package@\texttt{#1} package}\index{packages!#1@\texttt{#1}}% package name
    }%
}% command name -- adds backslash automatically
\newcommand{\docopt}[1]{\ensuremath{\langle}\textrm{\textit{#1}}\ensuremath{\rangle}}% optional command argument
\newcommand{\docarg}[1]{\textrm{\textit{#1}}}% (required) command argument
\newenvironment{docspec}{\begin{quotation}\ttfamily\parskip0pt\parindent0pt\ignorespaces}{\end{quotation}}% command specification environment
\newcommand{\docenv}[1]{\texttt{#1}\index{#1 environment@\texttt{#1} environment}\index{environments!#1@\texttt{#1}}}% environment name
\newcommand{\docenvdef}[1]{\hlred{\texttt{#1}}\label{env:#1}\index{#1 environment@\texttt{#1} environment}\index{environments!#1@\texttt{#1}}}% environment name
\newcommand{\docpkg}[1]{\texttt{#1}\index{#1 package@\texttt{#1} package}\index{packages!#1@\texttt{#1}}}% package name
\newcommand{\doccls}[1]{\texttt{#1}}% document class name
\newcommand{\docclsopt}[1]{\texttt{#1}\index{#1 class option@\texttt{#1} class option}\index{class options!#1@\texttt{#1}}}% document class option name
\newcommand{\docclsoptdef}[1]{\hlred{\texttt{#1}}\label{clsopt:#1}\index{#1 class option@\texttt{#1} class option}\index{class options!#1@\texttt{#1}}}% document class option name defined
\newcommand{\docmsg}[2]{\bigskip\begin{fullwidth}\noindent\ttfamily#1\end{fullwidth}\medskip\par\noindent#2}
\newcommand{\docfilehook}[2]{\texttt{#1}\index{file hooks!#2}\index{#1@\texttt{#1}}}
\newcommand{\doccounter}[1]{\texttt{#1}\index{#1 counter@\texttt{#1} counter}}

\newcommand{\studyq}[1]{\marginnote{Q: #1}}

\hypersetup{colorlinks}% uncomment this line if you prefer colored hyperlinks (e.g., for onscreen viewing)

% Generates the index
\usepackage{makeidx}
\makeindex

\setcounter{tocdepth}{3}
\setcounter{secnumdepth}{3}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% custom commands

\newtheorem{theorem}{\color{pastel-blue}Theorem}[section]
\newtheorem{lemma}[theorem]{\color{pastel-blue}Lemma}
\newtheorem{proposition}[theorem]{\color{pastel-blue}Proposition}
\newtheorem{corollary}[theorem]{\color{pastel-blue}Corollary}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{example}[1][Example]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{remark}[1][Remark]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\hyphenpenalty=5000

% more pastel ones
\xdefinecolor{pastel-red}{rgb}{0.77,0.31,0.32}
\xdefinecolor{pastel-green}{rgb}{0.33,0.66,0.41}
\definecolor{pastel-blue}{rgb}{0.30,0.45,0.69} % crayola blue
\definecolor{gray}{rgb}{0.2,0.2,0.2} % dark gray

\xdefinecolor{orange}{rgb}{1,0.45,0}
\xdefinecolor{green}{rgb}{0,0.35,0}
\definecolor{blue}{rgb}{0.12,0.46,0.99} % crayola blue
\definecolor{gray}{rgb}{0.2,0.2,0.2} % dark gray

\xdefinecolor{cerulean}{rgb}{0.01,0.48,0.65}
\xdefinecolor{ust-blue}{rgb}{0,0.20,0.47}
\xdefinecolor{ust-mustard}{rgb}{0.67,0.52,0.13}

%\newcommand\comment[1]{{\color{red}#1}}

\newcommand{\dy}{\partial}
\newcommand{\ddy}[2]{\frac{\dy#1}{\dy#2}}

\newcommand{\ab}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\cb}{\boldsymbol{c}}
\newcommand{\db}{\boldsymbol{d}}
\newcommand{\eb}{\boldsymbol{e}}
\newcommand{\lb}{\boldsymbol{l}}
\newcommand{\nb}{\boldsymbol{n}}
\newcommand{\tb}{\boldsymbol{t}}
\newcommand{\ub}{\boldsymbol{u}}
\newcommand{\vb}{\boldsymbol{v}}
\newcommand{\xb}{\boldsymbol{x}}
\newcommand{\wb}{\boldsymbol{w}}
\newcommand{\yb}{\boldsymbol{y}}

\newcommand{\Xb}{\boldsymbol{X}}

\newcommand{\ex}{\mathrm{e}}
\newcommand{\zi}{{\rm i}}

\newcommand\Real{\mbox{Re}} % cf plain TeX's \Re and Reynolds number
\newcommand\Imag{\mbox{Im}} % cf plain TeX's \Im

\newcommand{\zbar}{{\overline{z}}}

\newcommand\Def[1]{\textbf{#1}}

\newcommand{\qed}{\hfill$\blacksquare$}
\newcommand{\qedwhite}{\hfill \ensuremath{\Box}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% some extra formatting (hacked from Patrick Farrell's notes)
%  https://courses.maths.ox.ac.uk/node/view_material/4915
%

% chapter format
\titleformat{\chapter}%
  {\huge\rmfamily\itshape\color{pastel-red}}% format applied to label+text
  {\llap{\colorbox{pastel-red}{\parbox{1.5cm}{\hfill\itshape\huge\color{white}\thechapter}}}}% label
  {1em}% horizontal separation between label and title body
  {}% before the title body
  []% after the title body

% section format
\titleformat{\section}%
  {\normalfont\Large\itshape\color{pastel-green}}% format applied to label+text
  {\llap{\colorbox{pastel-green}{\parbox{1.5cm}{\hfill\color{white}\thesection}}}}% label
  {1em}% horizontal separation between label and title body
  {}% before the title body
  []% after the title body

% subsection format
\titleformat{\subsection}%
  {\normalfont\large\itshape\color{pastel-blue}}% format applied to label+text
  {\llap{\colorbox{pastel-blue}{\parbox{1.5cm}{\hfill\color{white}\thesubsection}}}}% label
  {1em}% horizontal separation between label and title body
  {}% before the title body
  []% after the title body

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

% Front matter
%\frontmatter

% r.3 full title page
%\maketitle

% v.4 copyright page

\chapter*{}

\begin{fullwidth}

\par \begin{center}{\Huge Statistical Mechanics 3/4H}\end{center}

\vspace*{5mm}

\par \begin{center}{\Large typed up by B. S. H. Mithrandir}\end{center}

\vspace*{5mm}

\begin{itemize}
  \item \textit{Last compiled: \monthyear}
  \item Adapted from notes of Veronika Hubeny and someone else (maybe Mukund?), Durham
  \item This was part of the Durham 3/4H elective. Includes introduction to
  classical thermodynamics and some quantum systems.
  \item It is sometimes said that there are probably two ways to approach
  thermodynamics: start from classical thermodynamics first, or start from a
  stat mech approach. These notes take the latter approach. From a physical
  understanding point of view, these particular notes might be more informative
  / useful if the intended user already has some exposure already to the
  traditional way of approaching thermodynamics.
  \item[]
  \item \TODO diagrams
  \item \TODO to possibly merge in some other notes (e.g. expand a bit more on
  classical thermodynamics; or maybe not and keep it separate to highlight
  differences in approach)
\end{itemize}

\par

\par Licensed under the Apache License, Version 2.0 (the ``License''); you may not
use this file except in compliance with the License. You may obtain a copy
of the License at \url{http://www.apache.org/licenses/LICENSE-2.0}. Unless
required by applicable law or agreed to in writing, software distributed
under the License is distributed on an \smallcaps{``AS IS'' BASIS, WITHOUT
WARRANTIES OR CONDITIONS OF ANY KIND}, either express or implied. See the
License for the specific language governing permissions and limitations
under the License.
\end{fullwidth}

%===============================================================================

\chapter{Classical thermodynamics}

The aim is to describe thermodynamics in \Def{equilibrium}.

%-------------------------------------------------------------------------------

\subsection{Postulates and definitions}

\Def{Extensive variables} $X_i$ of a thermodynamic system are those that depend
and scales proportionally to the system size. The set of all extensive variables
are sufficient to describe the system in equilibrium. Examples include
\begin{itemize}
  \item energy
  \item volume
  \item charge
  \item number of particles
\end{itemize}

Thermodynamic system can either be \Def{isolated} or \Def{interacting} with an
external environment. Isolated systems to not exchange any of the extensive
parameters the environment.

\vspace*{3mm}

\Def{Thermodynamic postulate 1}:

\begin{center}\textit{There exists and equilibrium state.}\end{center}

\begin{proposition}[Generalised first law of thermodynamics]
At equilibrium,
\begin{equation}
  \frac{\mathrm{d}X_i}{\mathrm{d}t} = 0
\end{equation}
for all extensive variables $X_i$ in a system.
\end{proposition}

The above is usually phrased as\sidenote{Sign of $W$ depends on convention.
Positive sign here as energy transferring into system.}
\begin{equation}
  \delta U = \delta Q + \delta W,
\end{equation}
where $U$ is the \Def{internal energy}, $Q$ is the \Def{heat}, and $W$ is
\Def{work done}. Ther conservation of internal energy gives
\begin{equation}
  \mathrm{d}U = T\ \mathrm{d}S - P\ \mathrm{d}V + \phi_i q_i,
\end{equation}
where $S$ is the \Def{entropy} (see later), $T$ is temperature, $P$ is
\Def{pressure}, $V$ is \Def{volume}, and $\phi_i$ and $q_i$ denote some other
form of change. The exact definitions of some of these variables with respect to
the more familiar thermodynamic quantities will be provided later.

\vspace*{3mm}

\Def{Thermodynamic postulate 2}:

\begin{center}\textit{There exists a function $S$ called the \Def{entropy} for all thermodynamic systems such that}\end{center}
\textit{
\begin{enumerate}
  \item $S = S(X)$, $X = \{X_i\}$, and $S$ is itself extensive;
  \item $S \in C^{\infty}$ (usually)
  \item $S$ is maximised as a function of $X$ when the system is in thermal
  equilibrium.
\end{enumerate}
}

The two postulates constrain the allowed physical processes that go on within a
thermodynamic system.

\begin{example}
\marginnote{Superscripts denote system, subscripts denote variable within a system.}
Given two systems $\mathcal{S}^{1,2}$ with extensive parameters $X^{1,2}$
respectively, for $\mathcal{S} = \mathcal{S}^1 \cup \mathcal{S}^2$, the total
set of extensive parameters is $X = X^1 + X^2$. Since entropy is maximised,
$S(X) = S(X^1 + X^2) \geq S_1(X^1) + S_2(X^2)$.

Suppose $S(X) = S_1(X^1 + \delta X^1) + S_2(X^2 + \delta X^2)$, where $\delta
X^2$ is the required change to bring the system total system to equilibrium.
Then we want $\delta X^! + \delta X^2 = 0$. Considering the variation, we have
\begin{equation*}
  \delta S = \frac{\partial S_1}{\partial X^1} \delta X^1 + \frac{\partial S_2}{\partial X^2} \delta X^2 = \delta X^1 \left(\frac{\partial S_1}{\partial X^1} - \frac{\partial S_2}{\partial X^2}\right).
\end{equation*}
Since $\delta S \geq 0$, the sign of $\delta X^1$ is the same as $(\partial S_1
/ \partial X^1 - \partial S_2 / \partial X^2)$, thus determining the direction
of flow for $X$.
\end{example}

Some properties of entropy:
\begin{itemize}
  \item $S(X)$ is a homogeneous function of degree 1, i.e. $S(\alpha X) =
  \alpha S(X)$ for some scalar $\alpha$;
  \item the defining relation of entropy in terms of other extensive parameters
  is called the \Def{entropy fundamental relation}, which provides a complete
  specific formulation of the thermodynamic system;
  \item physically, the entropy measures the amount of disorder in the system.
\end{itemize}

\Def{Intensive parameters} $Z_i$ are those that are \Def{conjugate} to the
extensive parameters $X_i$, i.e.
\begin{equation}
  Z_i = \frac{\partial S}{\partial X_i}.
\end{equation}
Clearly $Z_i$ has to be homogeneous functions of degree 0, and $Z_i(X)$ is
called the \Def{equation of state}, but requires additional information in order
to have a complete specification of the thermodynamic system.

\begin{lemma}
If two systems $\mathcal{S}^{1,2}$ interact and exchange extensive variables
$X^{1,2}$, at equilibrium, we have $Z^1 = Z^2$.
\end{lemma}
\begin{proof}
Since we are at equilibrium, we aim to maximise $S = S^1 + S^2$ keeping $X = X^1
+ X^2$ fixed. This can be done either by variation, with
\begin{equation*}
  0 = \delta S = \delta X^1 \left(\frac{\partial S_1}{\partial X^1} - \frac{\partial S_2}{\partial X^2}\right) = (Z^1 - Z^2) \delta X^1 \qquad \Rightarrow \qquad Z^1 = Z^2,
\end{equation*}
or by the method of Lagrange multiplies, we aim to maximise $S - \lambda(X^1 +
X^2)$ to get
\begin{equation*}
  \frac{\partial S}{\partial X^1} - \lambda = 0, \qquad \frac{\partial S}{\partial X^2} - \lambda = 0
\end{equation*}
which implies $\lambda = Z^1 = Z^2$.
\qed
\end{proof}

\vspace*{3mm}

\Def{Zeroth law of thermodynamics}\sidenote{More commonly, if two systems are in
thermal equilibrium with a third system, then they are in equlibrium with each
other. This a statement about \emph{transitivity} and establishes thermal
equilibrium as an equivalence relation from a mathematical point of view.}:

\begin{center}\textit{For two systems in thermal equilibrium with each other,
$Z^1_i = Z^2_i$ for all $i$.}\end{center}

When two systems $\mathcal{S}^{1,2}$ are not in equilibrium and $Z^1 > Z^2$,
$\mathrm{d}S > 0$ implies $X$ flows from $\mathcal{S}^1$ to $\mathcal{S}^2$, and
we have a \Def{directed flow}.

Instead of entropy $S$, we can take the internal energy $U$ as the fundamental
variable\sidenote{And really is what would be the more traditional way of
introducing classical thermodynamics.}. We have the \Def{energetic fundamental
relations} with intensive parameters
\begin{equation}
  Y_i = \frac{\partial U}{\partial X_i}.
\end{equation}
From this, we note that
\begin{itemize}
  \item \Def{temperature} $T$ is conjugate to entropy $S$, i.e.
  \begin{equation}
    T = \frac{\partial U}{\partial S};
  \end{equation}
  \item \Def{pressure} $P$ is conjugate to volume $V$, i.e.
  \begin{equation}
    P = -\frac{\partial U}{\partial V};
  \end{equation}
  \item \Def{chemical potential} $\mu$ is conjugate to particle number $N$, i.e.
  \begin{equation}
    \mu = \frac{\partial U}{\partial N}.
  \end{equation}
\end{itemize}
All signs are by convention. Occasionally, we employ the \Def{inverse
temperature}\sidenote{Sometimes this is defined with a factor of the Boltzmann
constant $k_B$, i.e. $\beta = (k_B T)^{-1}$.} variable $\beta$ given by
\begin{equation}
  \beta = \frac{1}{T} = \frac{\partial S}{\partial U}.
\end{equation}

\begin{example}
An alternative phrasing of the first law of thermodynamics\sidenote{Actually the
more traditional way of stating the first law.} is given as energy conservation
(Einstein summation convention implied)
\begin{equation}
  \mathrm{d}U = T\ \mathrm{d}S + \frac{\partial U}{\partial X_i}\ \mathrm{d}X_i = T\ \mathrm{d}S + Y_i\ \mathrm{d}X_i.
\end{equation}
Then, by rephrasing in the conjugate variables, we have
\begin{align*}
  \mathrm{d}S &= \beta\ \mathrm{d}U + Z_i\ \mathrm{d}X_i \\
    &= \beta(T\ \mathrm{d}S + Y_i\ \mathrm{d}X_i) + Z_i\ \mathrm{d}X_i \\
    &= \mathrm{d}S + (\beta Y_i + Z_i)\ \mathrm{d}X_i,
\end{align*}
implying
\begin{equation}
  Y_i = -T Z_i,
\end{equation}
i.e. all individual intensive parameters associated with $T$ and $S$ are related
by a factor of temperature.
\end{example}

A system is said to undergo a \Def{quasi-static} process if, at every instant of
time along the path in the phase space, one can approximate the system as being
in equilibrium. A quasi-static process is \Def{adiabatic} if the system does not
exchange heat with its surroundings, i.e. $\mathrm{d}S = 0$ along the
appropriate curve in phase space.

%-------------------------------------------------------------------------------

\subsection{Some examples}

\begin{enumerate}
  \item Consider an \Def{ideal gas}\sidenote{Randomly moving point particles
  that are not subject to inter-particle interactions.} in equilibrium, which is
  made to undergo a quasi-static expansion from $V_i$ to $V_f$ adiabatically.
  During this process, $PV^\gamma = \textnormal{constant}$, where $\gamma$ is
  some number (the \Def{adaibatic index} or the \Def{heat capacity ratio}). We
  aim to calculate the change in internal energy of the gas, and the work done
  and heat absorbed by the gas when it goes from $(P_i, V_i)$ to $(P_f, V_f)$.

  Starting from the first law with $\mathrm{d}U = T\ \mathrm{d}S - P\
  \mathrm{d}V$, since the process is adiabatic, $\mathrm{d}S = 0$, so $\mathrm{d}U
  = - P\ \mathrm{d}V$. By assumption,
  \begin{equation*}
    PV^\gamma = P_i V_i^\gamma = C \qquad \Rightarrow \qquad P = \frac{P_i
    V_i^\gamma}{V^\gamma},
  \end{equation*}
  so integrating gives
  \begin{align*}
    \Delta U = U_f - U_i &= \int_{V_i}^{V_f} - \frac{P_i V_i^\gamma}{V^\gamma}\; \mathrm{d}V \\
      &= -P_i V_i^\gamma \left[\frac{V^{-\gamma + 1}}{-\gamma + 1}\right]^{V_f}_{V_i}\\
      &= \frac{P_f V_f - P_i V_i}{\gamma - 1}
  \end{align*}
  via using $P_i V_i^\gamma = P_f V_f^\gamma$.

  \item From $S \sim (UVN)^{1/3}$, derive the equation of state.

  This is asking for the intensive variables and their relations to the
  extensive variables. For some constant $c$, we have $U = cS^3 / (VN)$, so
  \begin{equation*}
    T = \frac{\partial U}{\partial S} = \frac{3cS^2}{VN}, \quad
    P = -\frac{\partial U}{\partial V} = \frac{cS^3}{V^2 N}, \quad
    \mu = \frac{\partial U}{\partial N} = -\frac{cS^3}{VN^2}.
  \end{equation*}

  \item Derive the \Def{Euler relation}
  \begin{equation}
    U(S, X_i) = TS + X_i Y_i,
  \end{equation}
  where $Y_i$ are the intensive variables associated with $X_i$ via the internal
  energy (instead of a entropy).

  Since $U$ is extensive, we have $U(\lambda S, \lambda X_i) = \lambda U(S,
  X_i)$ by homogeneity, so by chain rule and taking a derivative with respect to
  $\lambda$,
  \begin{equation*}
    \frac{\partial U}{\partial (\lambda S)}(\lambda S, \lambda X_i) \frac{\partial (\lambda S)}{\partial \lambda} + \frac{\partial U}{\partial (\lambda X_i)}(\lambda S, \lambda X_i) \frac{\partial (\lambda X_i)}{\partial \lambda} = U(S, X_i).
  \end{equation*}
  If we set $\lambda = 1$, then the result follows on noting $\partial U /
  \partial S = T$, and $\partial U / \partial X_i = Y_i$.

  \item Prove the \Def{Gibbs--Duhem relation}
  \begin{equation}
    S\ \mathrm{d}T + X_i\ \mathrm{d}Y_i = 0.
  \end{equation}

  Differentiating the Euler relation leads to
  \begin{equation*}
    \mathrm{d}U = T\ \mathrm{d}S + S\ \mathrm{d}T + X_i\ \mathrm{d}Y_i + Y_i\ \mathrm{d}X_i,
  \end{equation*}
  and since the first law tells us $\mathrm{d}U = T\ \mathrm{d}S + Y_i\
\mathrm{d}X_i$, the result follows.

  \item Show that the Euler and Gibbs--Duhem relations can be written as
  \begin{equation*}
    S = \frac{1}{T} U + Z_i X_i, \qquad U\ \mathrm{d}\left(\frac{1}{T}\right) + X_i\ \mathrm{d}Z_i = 0,
  \end{equation*}
  where $Z_i$ is the intensive variables associated with the entropy.
  
  Recall that $Y_i = -T Z_i$ and $\mathrm{d}U = T\ \mathrm{d}S + Y_i\
  \mathrm{d}X_i$, which implies the first relation via substitution and
  integration. Differentiating the first relation leads to
  \begin{equation*}
    \mathrm{d}S = U\ \mathrm{d}\left(\frac{1}{T}\right) + \frac{1}{T}\ \mathrm{d}U + X_i\ \mathrm{d}Z_i + Z_i\ \mathrm{d}Z_i,
  \end{equation*}
  and upon using the first law by substituting $\mathrm{d}S$ leads to the second
  relation.
  
  \item For an ideal gas, given $PV = nRT$ and $U = (3/2)RNT$, where $R$ is the
  universal gas constant, derive the associated \Def{entropic fundamental
  relation} $S(U,V,N)$.

  To make the equation more exact, consider scaling by $N$ as $u = U/N$, $v =
  V/N$, which leads to
  \begin{equation*}
    Pv = RT, \qquad u = \frac{3}{2}RT.
  \end{equation*}
  Note that we have
  \begin{itemize}
    \item $X_i$: $V$, $N$;
    \item $Y_i$: $-P$, $\mu$,
    \item $Z_i$: $P/T$, $-\mu / T$.
  \end{itemize}
  Using the Gibbs--Duhem relation and scaling by $N$ leads to
  \begin{equation*}
    u\ \mathrm{d}\left(\frac{1}{T}\right) + v\ \mathrm{d}\left(\frac{P}{T}\right) = \mathrm{d}\left(\frac{\mu}{T}\right).
  \end{equation*}
  By computing the relevant derivatives, we have
  \begin{equation*}
    \mathrm{d}\left(\frac{\mu}{T}\right) = -\frac{3R}{2}\frac{\mathrm{d}u}{u} - R\frac{\mathrm{d}v}{v},
  \end{equation*}
  which upon integrating leads to
  \begin{equation*}
    \frac{\mu}{T} = \frac{3R}{2} \log\frac{u}{u_0} - R \log\frac{v}{v_0}.
  \end{equation*}
  Using Euler's relation in the form
  \begin{equation*}
    S = \frac{1}{T}U + \frac{P}{T}V - \frac{\mu}{T}N,
  \end{equation*}
  substituting accordingly results in
  \begin{equation*}
    S(U,V,N) = \frac{5}{2}NR + NR \log\left(\frac{u^{3/2}v}{u_0^{3/2}v}\right).
  \end{equation*}
\end{enumerate}

%-------------------------------------------------------------------------------

\subsection{Thermodynamic potentials}

Recall that the Legendre transform of a given function $f(x,y)$ is achieved via
the consideration of
\begin{equation*}
  \mathrm{d}f = \frac{\partial f}{\partial x}\mathrm{d}x + \frac{\partial f}{\partial y}\mathrm{d}y \equiv u(x,y)\ \mathrm{d}x + v(x,y)\ \mathrm{d}y.
\end{equation*}
To define $f$ as a function of $u$ rather than $x$, consider
\begin{equation*}
  g(v,y) = f - ux,
\end{equation*}
which then implies
\begin{equation*}
  \mathrm{d}y = \mathrm{d}f - u\ \mathrm{d}x - x\ \mathrm{d}u = v\ \mathrm{d}y - x\ \mathrm{d}u,
\end{equation*}
where $v = \partial g / \partial y$ and $x - \partial g / \partial u$.

The Legendre transform is an involute transformation of real valued convex
functions in one dimension, such that their derivatives of the original and
transformed functions are inverses of each other. Physically is used to convert
functions of one quantity to a function of its conjugate variable (e.g. position
to momenta, pressure to volume, temperature to entropy, and vice-versa). The
\Def{thermodynamic potentials} are Legendre transforms of $U$ with respect to
the various extensive parameters. Some important ones are:
\begin{itemize}
  \item \textbf{Helmholtz free energy}, which is the transformation of $U$
  with respect to $S$, given by
  \begin{equation}
    F(T, X_i) = U - T.
  \end{equation}
  
  \item \textbf{Enthalpy}, which is the transformation of $U$
  with respect to $V$, given by
  \begin{equation}
    H(P, S, X_i) = U + PV.
  \end{equation}
  
  \item \textbf{Gibbs free energy}, which is the transformation of $U$
  with respect to $S$ and $V$, given by
  \begin{equation}
    G(T, X_i) = U - TS + PV.
  \end{equation}
\end{itemize}
\marginnote{These could be thought of as the capacity to do mechanical and
non-mechanical work ($F$), to do non-mechanical work and release heat ($H$),
capacity to do non-mechanical (reversible) work ($G$), and capacity to do work
and release heat $U$.} Sometimes it is easier to consider these potentials
instead of directly trying to find the measurements for the extensive
parameters. Which one we choose depends on how the system is described. For
example, if it is natural to vary $T$ for example, then we consider $F$.

\begin{example}
Show that $\mathrm{d}F = -S\ \mathrm{d}T + Y_i\ \mathrm{d}X_i$.

From the definition of the Helmholtz free energy,
\begin{align*}
  \mathrm{d}F &= \mathrm{d}U - T\ \mathrm{d}S - S\ \mathrm{d}T\\
    &= (T\ \mathrm{d}S + Y_i\ \mathrm{d}X_i) - T\ \mathrm{d}S - S\ \mathrm{d}T\\
    &= -S\ \mathrm{d}T + Y_i\ \mathrm{d}X_i.
\end{align*}

Since $\mathrm{d}F = (\partial F / \partial T)\ \mathrm{d}T + (\partial F /
\partial X_i)\ \mathrm{d}X_i$, we note that
\begin{equation}
  -S(T, X_i) = \frac{\partial F}{\partial T}, \qquad Y_i(T, X_i) = \frac{\partial F}{\partial X_i},
\end{equation}
and similar identities hold for the other thermodynamic potentials, i.e., from
the thermodynamic potentials you can derive essentially all the things you want
to know.
\end{example}

%-------------------------------------------------------------------------------

\subsection{Maxwell relations}

For simplicity, we concentrate on a system with extensive parameters $U$, $S$,
$V$ and $N$.

%-------------------------------------------------------------------------------

\subsection{Phase transitions}

%===============================================================================

\chapter{Ensembles}

%-------------------------------------------------------------------------------

\subsection{Recap on probability and classical mechanics}

%-------------------------------------------------------------------------------

\subsection{Microcanonical ensemble}

%-------------------------------------------------------------------------------

\subsection{Canonical ensemble}

%-------------------------------------------------------------------------------

\subsection{Ideal gas example}

%===============================================================================

\chapter{Quantum statistics}

%===============================================================================



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% r.5 contents
%\tableofcontents

%\listoffigures

%\listoftables

% r.7 dedication
%\cleardoublepage
%~\vfill
%\begin{doublespace}
%\noindent\fontsize{18}{22}\selectfont\itshape
%\nohyphenation
%Dedicated to those who appreciate \LaTeX{} 
%and the work of \mbox{Edward R.~Tufte} 
%and \mbox{Donald E.~Knuth}.
%\end{doublespace}
%\vfill

% r.9 introduction
% \cleardoublepage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% actual useful crap (normal chapters)
\mainmatter

%\part{Basics (?)}


%\backmatter

%\bibliography{refs}
\bibliographystyle{plainnat}

%\printindex

\end{document}

